---
title: "02a-flowEMMI on Z project"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: github_document
params:
  name: "02a-flowEMMI on Z project" # change if you rename file
---

```{r here, message=FALSE,echo=F}
here::i_am(paste0(params$name, ".Rmd"), uuid = "278690ab-b570-439d-af41-1eaf8d025f11")
```

The purpose of this document is ...

```{r packages,message=F}
library("conflicted")
library(purrr)
library(dplyr)
library(mvtnorm)
library(flowEMMi)
library(flowCore)
library(flowWorkspace)
library(ggcyto)
library(tidyverse)
library(RColorBrewer)
library(knitr)
library(ellipse)
library(ggforce)
```

```{r directories,echo=F}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Z-project

### Import preprocessed data

```{r import,message=FALSE}
DAPI <- readRDS("~/MSc_new_data/new_DAPI.rds")

FDA_PI <- readRDS("~/MSc_new_data/new_FDA_PI.rds")

gating_DAPI<- readRDS("~/MSc_new_data/z_gating_list.rds")
```

### flowEMMI gating on DAPI

Based on the scatter plot in 01-data, we could tell that except for the sample taken in the surrounding region, the cells of other region samples are mostly concentrated in the range of 20000 to 60000.

Therefore, we will run the automated gating on surrounding region and other regions respectively.

-   Setting range for other regions: 20000-50000

-   Setting range for Surrounding region: 1000-50000

```{r flowEMMI-DAPI,results='hide',eval=F}
set.seed(1)
location <- c("Inner_zone","Middle_zone","Outer_zone","Whole_colony","Surrounding")
gating_DAPI <- list()

for (i in 1:4){
  data_name <- paste0(location[i],"_DAPI.fcs")
  data <- DAPI[[i]]
  fdo <- mkFlowDataObject(data, xChannel="PMT.1", yChannel="PMT.9")
  gating <- flowEMMi( fdo=fdo, xMin=20000, xMax=50000, yMin=20000, yMax=50000
                      , initFraction=0.01
                      , finalFraction=1.0
                      , minClusters=5, maxClusters=15, clusterbracket=2
                      , numberOfInits=5
                      , verbose=TRUE
                      , parallel=FALSE
                      , convergenceEpsilon=0.01
                      , whenToRemoveOverlaps = 20
                      , mergeWhenCenter = FALSE
                      , mergeWhenTwoCenters = FALSE
                      , thresholdForDeletion = 0.2
                      , threshold = 0.9
                      , considerWeights=TRUE
                      , plot = FALSE
                      , alpha=0.9
                      , minMinor=500)
  gating_DAPI[[i]] <- gating$best
}

# DAPI surrounding
DAPI_sur <- DAPI[["Surrounding_DAPI.fcs"]] 
fdo_sur <- mkFlowDataObject(DAPI_sur, xChannel="PMT.1", yChannel="PMT.9")
gating_sur <- flowEMMi( fdo=fdo_sur, xMin=1000, xMax=50000, yMin=1000, yMax=50000
                      , initFraction=0.01
                      , finalFraction=1.0
                      , minClusters=5, maxClusters=15, clusterbracket=2
                      , numberOfInits=5
                      , verbose=TRUE
                      , parallel=FALSE
                      , convergenceEpsilon=0.01
                      , whenToRemoveOverlaps = 20
                      , mergeWhenCenter = FALSE
                      , mergeWhenTwoCenters = FALSE
                      , thresholdForDeletion = 0.2
                      , threshold = 0.9
                      , considerWeights=TRUE
                      , plot = FALSE
                      , alpha=0.9
                      , minMinor=500)
gating_DAPI[[5]] <- gating_sur$best

```

### Gating plots on DAPI
```{r plot-DAPI,message=FALSE}
location <- c("Inner_zone","Middle_zone","Outer_zone","Whole_colony","Surrounding")
gating_DAPI_plot <- list()

for (i in 1:5){
  data_name <- paste0(location[i],"_DAPI.fcs")
  data <- DAPI[[i]]
  plots <- plotDensityAndEllipses(fcsData = data, ch1="PMT.1", ch2="PMT.9", alpha=0.9,
                            logScale = F, results = gating_DAPI[[i]],
                            title = data_name, plotRelevance = T,
                            ellipseDotSize = 0.5, axis_size=10, axisLabeling_size=10,
                            xlab = "Forward Scatter", ylab = "DAPI", font = "Arial")
  gating_DAPI_plot[[i]] <- plots$plot
}
```

In the above graphs, the ellipses are colored according to their cluster probability, high probability in red, medium probability in white, low probability in blue.

### Mahalanobis distance
```{r maha-DAPI,message=FALSE}
maha_result <- list()

for (x in 1:5){
  data <- DAPI[[x]]@exprs[,c(11,27)]
  
  mu <- split(gating_DAPI[[x]]@mu,seq_len(ncol(gating_DAPI[[x]]@mu)))
  sigma <- gating_DAPI[[x]]@sigma
  
  n_cells <- nrow(data)
  n_cluster <- length(sigma)
  maha_data <- matrix(NA,nrow=n_cells,ncol=n_cluster)
  
  for (i in 1:n_cells){
    for (j in 1:n_cluster){
      maha_data[i,j] <- mahalanobis(data[i,],mu[[j]],sigma[[j]])
    }
  }
  maha_result[[x]] <- maha_data
}


# rename distance matrix
for (i in 1:5){
  maha1 <- maha_result[[i]]
  coln <- ncol(maha1)
  maha <- maha1[,2:coln] %>% as.data.frame()
  
  # set 50 as distance cutoff
  for (j in 1:nrow(maha)){
    rv <- maha[j,1:ncol(maha)-1]
    if(all(rv>50)) { maha$Min[j] <- NA}
    else {maha$Min[j] <- which.min(rv)}
  }

  maha_result[[i]] <- maha
  
  table <-table(maha[,ncol(maha)]) %>% 
    kable(caption = paste0(location[i],"_DAPI"),
          col.names = c("No. of Cluster","Number of Cells"))
  
  print(table)
}
```

```{r maha-plot}
# plot
DAPI_maha_plot <- list()

for (i in 1:5){
  data1 <- DAPI[[i]]@exprs[,c(11,27)] 
  data2 <- maha_result[[i]]
  data <- cbind(data1,data2$Min) %>% as.data.frame()
  colnames(data)[3] <- "Cluster"
  data$Cluster <- as.factor(data$Cluster)
  
  #ellipses generated by 
  plot1 <- ggplot(data,aes(x=PMT.1,y=PMT.9))+
    geom_point(aes(color=Cluster))+geom_mark_ellipse(aes(color=Cluster))+
    ggtitle(paste0(location[i],"_DAPI"),subtitle = "Auto-generated ellipses")
  
  print(plot1)
  
  #ellipses from flowEMMI result
  ellipses <- gating_DAPI[[i]]
  num_ellipse <- length(ellipses@sigma)
  
  p <- plot(NULL,type="n",xlim=c(0,70000),ylim=c(0,70000),xlab="PMT.1",ylab="PMT.9")
  
  for (j in 2:num_ellipse){
    mu <- ellipses@mu[,j]
    sigma <- ellipses@sigma[[j]]
    eli <- ellipse::ellipse(centre=mu,x=sigma,level=0.95,npoints=200) 
    p <- p+lines(eli,type="l",col=j)
  }
}
```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
