---
title: "02a-flowEMMi for-Z-project"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: github_document
params:
  name: "02a-flowEMMi-for-Z-project" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "11675b32-9913-442e-9b4a-03cdc39afb65")
```

The purpose of this document is ...

```{r packages,message=FALSE}
library("conflicted")
library(purrr)
library(dplyr)
library(mvtnorm)
library(flowEMMi)
library(flowCore)
library(ggcyto)
library(tidyverse)
library(RColorBrewer)
library(knitr)
library(ellipse)
library(ggforce)
library(ggplot2)
library(ggpubr)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Import preprocessed data

```{r import,message=FALSE}
DAPI <- readRDS("~/Desktop/MSc_new_data/new_DAPI.rds")
FDA_PI <- readRDS("~/Desktop/MSc_new_data/new_FDA_PI.rds")
```

## DAPI

Based on the scatter plot in 01-data, after removing the debris, the channels for automated gating are between 200 and 65535. Therefore, we will run the flowEMMI v2 at [0,70000].

This is a complete code on how to perform clustering using flowEMMI. You can obtain gating results for interested variables by changing the "channel".

- PMT.1 stands for FSC, reveals cell size and shape

- PMT.2 stands for SSC, reveals intracellular complexity

- PMT.3 stands for FDA_PI stain

- PMT.9 stands for DAPI stain


### flowEMMI gating on individual region

For the samples dyed by DAPI, we would like to mainly focus on the gating results based on FSC and DAPI, namely PMT.1 and PMT.9.


```{r flowEMMI-DAPI,results='hide',eval=F}
set.seed(1)
location <- c("Inner_zone","Middle_zone","Outer_zone","Surrounding","Whole_colony")
gating_DAPI <- list()

for (i in 1:5){
  data_name <- paste0(location[i],"_DAPI.fcs")
  data <- DAPI[[i]]
  fdo <- mkFlowDataObject(data, xChannel="PMT.1", yChannel="PMT.9")
  gating <- flowEMMi( fdo=fdo, xMin=0, xMax=70000, yMin=0, yMax=70000
                      , initFraction=0.01
                      , finalFraction=1.0
                      , minClusters=5, maxClusters=15, clusterbracket=2
                      , numberOfInits=5
                      , verbose=FALSE
                      , parallel=FALSE
                      , convergenceEpsilon=0.01
                      , whenToRemoveOverlaps = 20
                      , mergeWhenCenter = FALSE
                      , mergeWhenTwoCenters = FALSE
                      , thresholdForDeletion = 0.2
                      , threshold = 0.9
                      , considerWeights=TRUE
                      , plot = FALSE
                      , alpha=0.9
                      , minMinor=500)
  gating_DAPI[[data_name]] <- gating$best
}
```

```{r,echo=FALSE}
gating_DAPI <- readRDS("~/Desktop/MSc_new_data/gating_DAPI.rds")
gating_DAPI12 <- readRDS("~/Desktop/MSc_new_data/gating_DAPI12.rds")
gating_DAPI29 <- readRDS("~/Desktop/MSc_new_data/gating_DAPI29.rds")
```

### Gating plots

```{r plot-DAPI,message=FALSE}
gating_DAPI_plot <- list()

for (i in 1:5){
  data_name <- names(DAPI)[i]
  data <- DAPI[[i]]
  plots <- plotDensityAndEllipses(fcsData = data, ch1="PMT.1", ch2="PMT.9", alpha=0.9,
                            logScale = F, results = gating_DAPI[[i]],
                            title = data_name, plotRelevance = T,gridsize = 1000,
                            ellipseDotSize = 0.5, axis_size=10, axisLabeling_size=10,
                            xlab = "Forward Scatter", ylab = "DAPI", font = "Arial")
  gating_DAPI_plot[[i]] <- plots$plot
}

```

In the above graphs, the ellipses are colored according to their cluster probability, high probability in red, medium probability in white, low probability in blue.

## Metaclustering

In order to get meta-clustering results, we merge all gating results regardless of regions, then run the flowEMMI v2 to remove the overlap ellipses.

```{r meta-DAPI}
#PMT.1--Forward Scatter--col11
#PMT.9--Fluor. DAPI--col27

all_gate <- gating_DAPI[[1]]
all_gate@mu <- all_gate@mu[,-1]
all_gate@sigma <- all_gate@sigma[-1]
all_gate@clusterProbs <- all_gate@clusterProbs[-1]

for (i in 2:5){
    mu <- gating_DAPI[[i]]@mu[,-1]
    sigma <- gating_DAPI[[i]]@sigma[-1]
    prob <- gating_DAPI[[i]]@clusterProbs[-1]
    all_gate@mu <- cbind(all_gate@mu,mu)
    all_gate@sigma <- c(all_gate@sigma,sigma)
    all_gate@clusterProbs <- c(all_gate@clusterProbs,prob)
}

Gating_DAPI_meta <- removeOverlaps(em=all_gate, alpha=0.9
                                  , mergeWhenCenter = FALSE
                                  , mergeWhenTwoCenters = FALSE
                                  , thresholdForDeletion = 0.9
                                  , threshold = 0.9
                                  , shrinkingFunction=shrinkEllipses
                                  , considerWeights=TRUE
                                  , plot = FALSE
                                  , minMinor = 500)
```

### Meta-gating plots

Now, we'd like to illustrate the meta-clustering effects on different regions.

```{r meta-DAPI-plot,message=FALSE}
gating_DAPI_meta_plot <- list()

for (i in 1:5){
  data_name <- names(DAPI)[i]
  data <- DAPI[[i]]
  plots <- plotDensityAndEllipses(fcsData = data, ch1="PMT.1", ch2="PMT.9", alpha=0.9,
                            logScale = F, results = Gating_DAPI_meta,
                            title = paste0("Meta-clustering on ",data_name), 
                            plotRelevance = T,gridsize = 1000,
                            ellipseDotSize = 0.5, axis_size=10, axisLabeling_size=10,
                            xlab = "Forward Scatter", ylab = "DAPI", font = "Arial")
  gating_DAPI_meta_plot[[i]] <- plots$plot
}

```

## Mahalanobis distance function with flowEMMi

Following is the integrated function.

Given a preprocessed dataset and the gating result from above, it can generate the:

-   Mahalanobis distance matrix & Cluster result

-   Parameter table: Number of cells in each cluster, Area of the clustering ellipse, and the Coordinates of center

-   Gating plots

```{r}
flowEMMi_mahalanobis <- function(data,data_name,gating_data,alpha){
  
  mu <- gating_data@mu
  sigma <- gating_data@sigma
  
  names <- colnames(data)
  
  n_cells <- nrow(data)
  n_clusters <- length(sigma)
  
  #generate mahalanobis matrix
  maha_data <- matrix(NA,nrow=n_cells,ncol=n_clusters)
  
  for (i in 1:n_cells){
    for(j in 1:n_clusters){
      maha_data[i,j] <- mahalanobis(data[i,],mu[,j],sigma[[j]])
    }
  }
  
  maha <- maha_data[,2:n_clusters] %>% as.data.frame()
  
  #set 95% quantile as cutoff value
  threshold <- -2*log(1-alpha)
  
  #determine cluster
  for (cell in 1:n_cells){
    rv <- maha[cell,1:n_clusters-1]
    if(all(rv>threshold)) { maha$Cluster[cell] <- NA}
    else { maha$Cluster[cell] <- which.min(rv)}
  }
  
  maha_data2 <- cbind(data,maha)
  
  test <- base::table(maha[,ncol(maha)]) %>% as.data.frame()
  coordinates <- sprintf("(%.2f,%.2f)",mu[1,2:ncol(mu)],mu[2,2:ncol(mu)])
  
  #Area of Ellipse
  eigen <- matrix(NA,nrow=length(sigma),ncol=2)
  for (i in 1:length(sigma)){
    eigen[i,] <- eigen(sigma[[i]])$values
  }
  eigen <- eigen[-1,]
  
  area <- matrix(NA,nrow=nrow(eigen),ncol=1)
  for (i in 1:nrow(eigen)){
    area[i,1] <- pi*sqrt(eigen[i,1]*eigen[i,2])*(-2*log(1-alpha))
  }
  
  area <- area %>% as.data.frame()
  test <- cbind(test,area,coordinates)
  colnames(test) <- c("Cluster","Cells","Area","Coordinate")

  #plot
  maha_data2$Cluster <- as.factor(maha_data2$Cluster)
  
  plot1 <- ggplot(maha_data2,aes(x=!!sym(names[1]),y=!!sym(names[2]),color=Cluster))+
    geom_point(alpha=1/5)+
    ggtitle(data_name)
  
  num_ellipse <- length(gating_data@sigma)
  
  for (j in 2:num_ellipse){
    mu <- gating_data@mu[,j]
    sigma <- gating_data@sigma[[j]]
    eli <- ellipse::ellipse(centre=mu,x=sigma,level=alpha,npoints=200) 
    eli <- as.data.frame(eli)
    colnames(eli)<- names
    plot1 <- plot1+geom_path(data = eli,
                  aes(x=!!sym(names[1]),y=!!sym(names[2])),color=j)
  }
  
  maha_result <- list(maha_matrix=maha_data2,table_info=test,plot=plot1)
  return(maha_result)
}
```

### Maha-dist tables and plots

Based on the above functions, we substitute preprocessed dataset and gating results to generate tables and images for different regions. The tables list the number of cells in each cluster and the size of the ellipses for each cluster. 

```{r maha-DAPI}
# FSC vs. DAPI
flowemmi_DAPI <- list()
#flowemmi_meta_DAPI <- list()

for (i in 1:5){
  data_name <- names(DAPI)[i]
  data <- DAPI[[i]]@exprs[,c(11,27)]
  gating_data <- gating_DAPI[[i]]
  flowemmi_DAPI[[data_name]] <-flowEMMi_mahalanobis(data,data_name,gating_data,0.9)
  
  #dname <- paste0("Meta_",data_name)
  #flowemmi_meta_DAPI[[data_name]] <- flowEMMi_mahalanobis(data,dname,Gating_DAPI_meta,0.9)
  
  # visualization
  table <- flowemmi_DAPI[[data_name]]$table_info %>% 
    kable(caption = data_name,
          col.names = c("Cluster","Cells","Area","Coordinate"))
  print(table)
  print(flowemmi_DAPI[[data_name]]$plot)
}
```


## ROC

In the following section, we will ues the ROC-like curves to reflect the relationship between the number of cells contained in each cluster and their corresponding ellipse sizes, involving comparisons between different sampling positions.


```{r roc}
flowroc <- function(data,gating_data,CI){
  data1 <- data[,-c(1,2,ncol(data))]
  ncells <- nrow(data1)
  
  roc.points <- data.frame(cell=NA,volume=NA)
  
  for (alpha in CI){
    threshold <- -2*log(1-alpha)
  
    for (cell in 1:ncells){
      rv <- data1[cell,1:ncol(data1)-1]
      if(all(rv>threshold)) {data1$Cluster[cell] <- NA}
      else {data1$Cluster[cell] <- which.min(rv)}
    }
  
    result <- base::table(data1$Cluster) %>% as.data.frame()
    sum1 <- sum(result$Freq)
  
    sigma <- gating_data@sigma
  
    eigen <- matrix(NA,nrow=length(sigma),ncol=2)
    for (i in 1:length(sigma)){
      eigen[i,] <- eigen(sigma[[i]])$values
    }
    eigen <- eigen[-1,]
  
    area <- matrix(NA,nrow=nrow(eigen),ncol=1)
    for (i in 1:nrow(eigen)){
      area[i,1] <- pi*sqrt(eigen[i,1]*eigen[i,2])*(-2*log(1-alpha))
    }
    area <- as.data.frame(area)
    sum2 <- sum(area)
    sum.roc <- cbind(cell=sum1,volume=sum2)
    
    roc.points <- rbind(roc.points,sum.roc)
  }
  return(roc.table=roc.points)
}
```


```{r,echo=F}
gating_DAPI12 <- readRDS("~/Desktop/MSc_new_data/gating_DAPI12.rds")
gating_DAPI29 <- readRDS("~/Desktop/MSc_new_data/gating_DAPI29.rds")
```

Calculate ellipse sizes and included cell numbers at different alpha levels.

```{r roc-DAPI,eval=FALSE}
roc.result <- list()
roc.meta.result <- list()

# FSC vs. DAPI
# using gating results on each region
for (i in 1:5){
  data_name <- names(DAPI)[i]
  data <- flowemmi_DAPI[[data_name]]$maha_matrix
  
  gating_data <- gating_DAPI[[data_name]]
  alpha <- seq(0.01,0.99,0.01)
  
  roc.result[[data_name]] <- flowroc(data,gating_data,alpha)
}

# using meta-gating on each region
for (i in 1:5){
  data_name <- names(DAPI)[i]
  data <- flowemmi_meta_DAPI[[data_name]]$maha_matrix
  alpha <- seq(0.01,0.99,0.01)
  
  roc.meta.result[[data_name]] <- flowroc(data,Gating_DAPI_meta,alpha)
}

```


After computing the sum of cells' number and the ellipse volume at different confidence intervals, we can now plot the ROC-like curves to compare the trade-off tendency for different locations.

```{r roc-plot}
# plot
plot_point <- function(roc_result){
  
  plot.points <- data.frame()
  plot.scale <- data.frame()
  plot.scale.points <- data.frame()

  for (i in 1:5){
    data <- roc_result[[i]]
    data[1,] <- c(0,0)
  
  # No Scaling
  data$location <- substr(names(DAPI)[i],1,nchar(names(DAPI)[i])-9) %>% as.factor()
  plot.points <- rbind(plot.points,data)
  
  # Scaled by regional maximum
  max <- data[100,]
  for (j in 1:100){
    data[j,1] <- data[j,1]/max[1,1]
    data[j,2] <- data[j,2]/max[1,2]
  }
  plot.scale <- rbind(plot.scale,data)
  
  
  # Scaled by the max number among all regions
  plot.scale.points <- plot.points
  max1 <- max(plot.scale.points[,1])
  max2 <- max(plot.scale.points[,2])
  
  for (j in 1:nrow(plot.points)){
    plot.scale.points[j,1]<- plot.scale.points[j,1]/max1
    plot.scale.points[j,2]<- plot.scale.points[j,2]/max2
   }
  }
  
  points <- list(raw.point=plot.points,regional.scaled=plot.scale,all.scaled=plot.scale.points)
  return(points)
}
```


```{r,echo=F}
roc_19 <- readRDS("~/Desktop/MSc_new_data/roc_DAPI.rds")
roc_29 <- readRDS("~/Desktop/MSc_new_data/roc_DAPI29.rds")
roc_12 <- readRDS("~/Desktop/MSc_new_data/roc_DAPI12.rds")

roc_meta_19 <- readRDS("~/Desktop/MSc_new_data/roc_DAPI_meta.rds")
```

### Different channels and regions

```{r DAPI-channel}
channels <- c("roc_12","roc_19","roc_29")
names<-c("FSC vs. SSC","FSC vs. DAPI","SSC vs. DAPI")
point <- list()

for (i in 1:3){
  data <- get(channels[i])
  roc <- plot_point(data)
  name <- names[i]
  
  point[[i]] <- ggplot(roc$raw.point,aes(x=volume,y=cell,color=location))+
    geom_line()+labs(x="volume",y="cell")+
    coord_cartesian(ylim = c(0,60000),xlim=c(0,2e9))+
    scale_x_continuous(breaks = seq(0,2e9,by=1e9),labels = c(0,1e9,2e9))+
    theme_minimal()
  
  point[[i+3]] <- ggplot(roc$regional.scaled,aes(x=volume,y=cell,color=location))+
    geom_line()+labs(x="volume",y="cell")+
    theme_minimal()
  
  point[[i+6]] <- ggplot(roc$all.scaled,aes(x=volume,y=cell,color=location))+
    geom_line()+labs(x="volume",y="cell")+
    theme_minimal()
}

complete <- ggarrange(plotlist = point,ncol=3, nrow=3, common.legend = TRUE, legend="bottom")
annotate_figure(complete,
                top = "           FSC vs. SSC                           FSC vs. DAPI                                 SSC vs. DAPI",
                left = "                 Global                     Regional           None Scaled")
```

### Gating vs. meta-clustering

```{r DAPI-meta}
location <- c("Inner_zone","Middle_zone","Outer_zone","Surrounding","Whole_colony")
point <- list()

p1 <- plot_point(roc_19)
p2 <- plot_point(roc_meta_19)

for (i in 1:5){
  var <- location[i]
  
  data1 <- p1$raw.point %>% dplyr::filter(location==var)
  data2 <- p2$raw.point %>% dplyr::filter(location==var)
  
  data1$Type <- "gating"
  data2$Type <- "meta-clustering"
  
  data <- rbind(data1,data2)
  
  point[[i]] <- ggplot(data,aes(x=volume,y=cell,color=Type))+
    geom_line()+labs(x="volume",y="cell",title=var)+
    coord_cartesian(ylim = c(0,65535),xlim=c(0,1.5e9))+
    scale_x_continuous(breaks = seq(0,1.5e9,by=1.5e9),labels = c(0,1.5e9))+
    theme_minimal()
}

complete <- ggarrange(plotlist = point,
          ncol=3, nrow=2, common.legend = TRUE, legend="bottom")
annotate_figure(complete,top = "FSC vs. DAPI: Without Scaling")
```

### Different methods
```{r DAPI-method}
roc_flowEMMI <- readRDS("~/Desktop/MSc_new_data/roc_DAPI.rds")
roc_GMM_flex <- readRDS("~/Desktop/MSc_new_data/DAPI_GMM_flex_roc.rds")
roc_GMM_fix <- readRDS("~/Desktop/MSc_new_data/DAPI_GMM_fix_roc.rds")
roc_fC_flex <- readRDS("~/Desktop/MSc_new_data/DAPI_fC_flex_roc.rds")
roc_fC_fix <- readRDS("~/Desktop/MSc_new_data/DAPI_fC_fix_roc.rds")

point <- list()

for (i in 1:5){
data1 <- roc_flowEMMI[[i]]
data1$method <- as.factor("flowEMMI")

data2 <- roc_GMM_flex[[i]]$roc.table
data2$method <- as.factor("GMM")

data3 <- roc_GMM_fix[[i]]$roc.table
data3$method <- as.factor("GMM_K")

data4 <- roc_fC_flex[[i]]$roc.table
data4$method <- as.factor("flowClust")

data5 <- roc_fC_fix[[i]]$roc.table
data5$method <- as.factor("flowClust_K")

plot.points <- data.frame()
plot.scale <- data.frame()
plot.scale.points <- data.frame()

for (m in 1:5){
  var <- paste0("data",m)
  data <- get(var)
  if (m<4) {data[1,1:2] <- c(0,0)}
  
  # No Scaling
  plot.points <- rbind(plot.points,data)
  
  # Scaled by regional maximum
  max=data[nrow(data),]
  
  for (j in 1:100){
    data[j,1] <- data[j,1]/max[1,1]
    data[j,2] <- data[j,2]/max[1,2]
  }
  plot.scale <- rbind(plot.scale,data)
  
  # Scaled by the max number among all regions
  plot.scale.points <- plot.points
  max1 <- max(plot.scale.points[,1])
  max2 <- max(plot.scale.points[,2])
  
  for (j in 2:nrow(plot.points)){
    plot.scale.points[j,1]<- plot.scale.points[j,1]/max1
    plot.scale.points[j,2]<- plot.scale.points[j,2]/max2
  }
  
  point[[i]] <- ggplot(plot.points, aes(x = volume, y = cell, color = method)) +
    geom_line() + 
    coord_cartesian(xlim=c(0,5e9),ylim=c(0,80000)) +
    scale_x_continuous(breaks = seq(0,4e9,by=4e9),labels = c(0,4e9))+
    labs(x = "volume", y = "cell") +
    theme_minimal()
  
  point[[i+5]] <- ggplot(plot.scale, aes(x = volume, y = cell, color = method)) +
    geom_line() +
    labs(x = "volume", y = "cell") +
    coord_cartesian(xlim=c(0,1),ylim=c(0,1)) +
    scale_x_continuous(breaks = seq(0,1,by=0.5),labels = c(0,0.5,1))+
    theme_minimal()
  
  point[[i+10]] <- ggplot(plot.scale.points, aes(x = volume, y = cell, color = method)) +
    geom_line() +
    labs(x = "volume", y = "cell") +
    scale_x_continuous(breaks = seq(0,1,by=0.5),labels = c(0,0.5,1))+
    theme_minimal()
  }
}

complete <- ggarrange(plotlist = point,ncol=5, nrow=3, common.legend = TRUE, legend="bottom")
annotate_figure(complete,
                top = "Inner_zone         Middle_zone       Outer_zone       Surrounding        Whole_colony",
                left = "            Global               Regional                None Scaled") 

```



## Files written

These files have been written to the target directory, `r paste0("data/", params$name)`:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
