---
title: "02b-flowEMMI on Z-project (FDA_PI)"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: github_document
params:
  name: "02b-flowEMMI on Z-project (FDA_PI)" # change if you rename file
---

```{r here, message=FALSE,echo=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "20fc4da1-1f73-434a-bd4c-2f3028f7f3ad")
```


```{r packages}
library("conflicted")
library(purrr)
library(dplyr)
library(mvtnorm)
library(flowEMMi)
library(flowCore)
library(flowWorkspace)
library(ggcyto)
library(tidyverse)
library(RColorBrewer)
library(knitr)
library(ellipse)
library(ggforce)
```

```{r directories,echo=FALSE}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Z-project 

### Import preprocessed data

```{r import,message=FALSE}
DAPI <- readRDS("~/MSc_new_data/new_DAPI.rds")
FDA_PI <- readRDS("~/MSc_new_data/new_FDA_PI.rds")

```

### flowEMMI gating on FDA_PI

Based on the scatter plot in 01-data, we would like to set range as below.

-   Setting range for regions: 2000-65000

```{r flowEMMI-FDAPI,results='hide'}
set.seed(1)
location <- c("Inner_zone","Middle_zone","Outer_zone","Surrounding","Whole_colony")
gating_FDA_PI <- list()

for (i in 1:5){
  data_name <- paste0(location[i],"_FDA_PI.fcs")
  data <- FDA_PI[[i]]
  fdo <- mkFlowDataObject(data, xChannel="PMT.1", yChannel="PMT.3")
  gating <- flowEMMi( fdo=fdo, xMin=2000, xMax=65000, yMin=2000, yMax=65000
                      , initFraction=0.01
                      , finalFraction=1.0
                      , minClusters=5, maxClusters=15, clusterbracket=2
                      , numberOfInits=5
                      , verbose=TRUE
                      , parallel=FALSE
                      , convergenceEpsilon=0.01
                      , whenToRemoveOverlaps = 20
                      , mergeWhenCenter = FALSE
                      , mergeWhenTwoCenters = FALSE
                      , thresholdForDeletion = 0.2
                      , threshold = 0.9
                      , considerWeights=TRUE
                      , plot = FALSE
                      , alpha=0.9
                      , minMinor=500)
  gating_FDA_PI[[i]] <- gating$best
}

```

### Gating plots on FDA_PI
```{r plot-FDAPI,message=FALSE}
location <- c("Inner_zone","Middle_zone","Outer_zone","Surrounding","Whole_colony")
gating_FDA_PI_plot <- list()

for (i in 1:5){
  data_name <- paste0(location[i],"_FDA_PI.fcs")
  data <- FDA_PI[[i]]
  plots <- plotDensityAndEllipses(fcsData = data, ch1="PMT.1", ch2="PMT.3", alpha=0.9,
                            logScale = F, results = gating_FDA_PI[[i]],
                            title = data_name, plotRelevance = T,
                            ellipseDotSize = 0.5, axis_size=10, axisLabeling_size=10,
                            xlab = "Forward Scatter", ylab = "DAPI", font = "Arial")
  gating_FDA_PI_plot[[i]] <- plots$plot
}
```

In the above graphs, the ellipses are colored according to their cluster probability, high probability in red, medium probability in white, low probability in blue.

### Mahalanobis distance
```{r maha-FDAPI,message=FALSE}
maha_result <- list()

for (x in 1:5){
  data <- FDA_PI[[x]]@exprs[,c(11,15)]
  
  mu <- split(gating_FDA_PI[[x]]@mu,seq_len(ncol(gating_FDA_PI[[x]]@mu)))
  sigma <- gating_FDA_PI[[x]]@sigma
  
  n_cells <- nrow(data)
  n_cluster <- length(sigma)
  maha_data <- matrix(NA,nrow=n_cells,ncol=n_cluster)
  
  for (i in 1:n_cells){
    for (j in 1:n_cluster){
      maha_data[i,j] <- mahalanobis(data[i,],mu[[j]],sigma[[j]])
    }
  }
  maha_result[[x]] <- maha_data
}


# rename distance matrix
for (i in 1:5){
  maha1 <- maha_result[[i]]
  coln <- ncol(maha1)
  maha <- maha1[,2:coln] %>% as.data.frame()
  
  # set 50 as distance cutoff
  for (j in 1:nrow(maha)){
    rv <- maha[j,1:ncol(maha)-1]
    if(all(rv>50)) { maha$Min[j] <- NA}
    else {maha$Min[j] <- which.min(rv)}
  }

  maha_result[[i]] <- maha
  
  table <-table(maha[,ncol(maha)]) %>% 
    kable(caption = paste0(location[i],"_FDA_PI"),
          col.names = c("No. of Cluster","Number of Cells"))
  
  print(table)
}
```

```{r maha-plot}
# plot
FDA_PI_maha_plot <- list()

for (i in 1:5){
  data1 <- FDA_PI[[i]]@exprs[,c(11,15)] 
  data2 <- maha_result[[i]]
  data <- cbind(data1,data2$Min) %>% as.data.frame()
  colnames(data)[3] <- "Cluster"
  data$Cluster <- as.factor(data$Cluster)
  
  #ellipses generated by 
  plot1 <- ggplot(data,aes(x=PMT.1,y=PMT.3))+
    geom_point(aes(color=Cluster))+geom_mark_ellipse(aes(color=Cluster))+
    ggtitle(paste0(location[i],"_FDA_PI"),subtitle = "Auto-generated ellipses")
  
  print(plot1)
  
  #ellipses from flowEMMI result
  ellipses <- gating_FDA_PI[[i]]
  num_ellipse <- length(ellipses@sigma)
  
  p <- plot(NULL,type="n",xlim=c(0,70000),ylim=c(0,70000),xlab="PMT.1",ylab="PMT.3")
  
  for (j in 2:num_ellipse){
    mu <- ellipses@mu[,j]
    sigma <- ellipses@sigma[[j]]
    eli <- ellipse::ellipse(centre=mu,x=sigma,level=0.95,npoints=200) 
    p <- p+lines(eli,type="l",col=j)
  }
}
```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
